---
title: "metabarcoding analysis example"
author: "Kimberly Ledger"
date: "2024-01-11"
output: html_document
---
## Load libraries and data 

load libraries 
```{r, message=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
```

load ASV table and metadata
```{r}
asv_table <- readRDS("/home/kimberly.ledger/AFSC_eDNA/pink_herring_diet/filtered.seqTab.RDS") %>%
  select(!Row.names)

#transpose 
asv_table <- data.frame(t(asv_table))

#set column names to be ASV# 
colnames(asv_table) <- asv_table["ASV",]

#remove extra row with ASV ID 
asv_table <- asv_table[-144,] 

#make sure reads are numbers
# Convert all character columns to numeric
for (col in names(asv_table)) {
  asv_table[[col]] <- as.numeric(asv_table[[col]])
}

#make sample ID a column 
asv_table$Sample_ID <- rownames(asv_table)
asv_table$Sample_ID <- as.factor(asv_table$Sample_ID)
```

### replace ASVs with taxonomic assignments

because the insect and the blastn classifiers aren't perfect on their own, i combined the outputs by hand for our remaining ASVs
```{r}
my_tax <- read.csv("pink_herring_diet/asv_taxonomy_pinkherringdiet_custom.csv")
```


join to taxononmy and group reads by taxon assigment - this removes ASVs without a taxonomic assignment 
```{r}
taxon_table <- asv_table %>%
  pivot_longer(cols = c(1:411), names_to = "ASV", values_to = "reads") %>%  ## change column numbers in asv_table minus one (for the Sample_ID)
  separate(ASV, into = c("ASV_label", "ASV_num"), remove = F) %>% ## this line and the next are only needed it your ASV IDs do not match (i.e. ASV0001 in asv_table and ASV1 in taxonomic table)
  mutate(ASV_num = as.integer(ASV_num)) %>%     
  left_join(my_tax, by = "ASV_num") %>%       ## otherwise, join by "ASV"
  select(!ASV_label) %>%
  select(!ASV_num) %>% 
  select(!representative) %>%
  group_by(Sample_ID, taxon, rank, order, family, genus, species) %>%
  summarize(reads = sum(reads)) %>%
  filter(!is.na(taxon))
```

### join metadata to asv table 

```{r}
metadata <- read.csv("/home/kimberly.ledger/AFSC_eDNA/pink_herring_diet/pinkherring_diet_metadata.csv") %>%
  dplyr::rename(Sample_ID = extraction_ID)

#illumina output changed "_" to "-"
metadata$Sample_ID <- gsub("_", "-", metadata$Sample_ID) 

taxon_table_with_metadata <- metadata %>%
  left_join(taxon_table, by = "Sample_ID")
```


### plot

let's start by visualizing the reads - absolute number of reads   
```{r}
taxon_table_with_metadata %>%
  #filter(sample_type == "positive") %>%  can filter dataframe here to plot less sampels at one time
  ggplot(aes(x=Sample_ID, y=reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
    theme_bw() +
  labs(
    y = "number of sequencing reads",
    x = "sample ID",
    title = "") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "none",  # too many asvs to plot entire 
    legend.title = element_blank()
  )
```

plot them - proportion 
```{r, fig.height=6}
 taxon_table_with_metadata %>%
  #filter(sample_type == "positive") %>%  can filter dataframe here to plot less sampels at one time 
  group_by(Sample_ID) %>%
  mutate(ReadsPerSample = sum(reads)) %>%
  mutate(read_prop = reads/ReadsPerSample) %>% 
  ggplot(aes(x=Sample_ID, y=read_prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
    theme_bw() + 
   labs(
    y = "proportion of sequencing reads",
    x = "sample ID",
    title = "")  +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.position = "right",
    legend.text = element_text(size = 6),
    legend.title = element_blank()
  )  
```

