---
title: "blastn_custom_formatting"
author: "Kimberly Ledger"
date: "2023-11-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-libraries}
library(dplyr)
library(tidyr)
library(stringr)
```

## load output from taxonkit 
```{r load-output-from-taxonkit}
taxonomy <-read.delim("~/AFSC_eDNA/crab_diet/crabdiet_18S/blastn_taxlineage.txt", header = FALSE, na.strings=c(""," ","NA"))

head(taxonomy)
```

## clean up the header a bit
```{r}
# use the full taxonomy rather than the seq id to collapse identical entries
tax_df <- taxonomy %>%
  filter(V4 > 100) %>% # make sure all retained matches are >100 bp
  select(-V2, -V5, -V6, -V7, -V14, -V8, -V9, -V10, -V11, -V12, -V15, -V16) %>%  #remove unnecessary columns
  group_by(V1, V17) %>% # group by the sequence key and the full taxonomy to reduce duplicate entries
  unique() %>% # doing that reduced the number of entries from 146k to 17k
  dplyr::rename(qseqid=V1, perc_id=V3, length=V4, sscinames=V13, taxonomy=V17) %>% #rename headers
  filter(!str_detect(taxonomy, "environmental")) %>% # filter out any environmental samples
  filter(!str_detect(taxonomy, "synthetic")) %>% # filter out any synthetic "samples"
  filter(!str_detect(taxonomy, "uncultured")) %>% # filter out any uncultured "samples"
  filter(!str_detect(taxonomy, "clone")) %>% # filter out any clone "samples"
  filter(!str_detect(taxonomy, "sp.")) %>% # remove anything with "sp."
  filter(!str_detect(taxonomy, "var."))
```

## calcualate the number of ASV's with taxonomic assignment
```{r how-many-total-ASVs-w-taxonomy}
tax_df %>% 
  ungroup() %>%
  select(qseqid) %>%
  unique() %>%
  nrow
```

since many crab OTU's don't have many reads or i will filter by the OTU's with data to make this dataset smaller... 

```{r}
keeper_otu <- readRDS("18S_OTUs_keep.RDS")
```

```{r}
tax_df <- tax_df %>%
  filter(qseqid %in% keeper_otu$ASV)
```


## reformat table 
```{r}
# formatting the taxonomy variables
taxon_df <- tax_df %>%
  filter(str_detect(taxonomy, ";")) %>%
  separate(taxonomy, into=c("kingdom", "phylum", "class", "order", "family", "genus", "species"), sep = ";") %>%
  filter(species != "Eukaryota sp.")
```



```{r}
taxon_df %>%
  filter(qseqid == "OTU_038")
```


# Sort and clean-up the results based on % identity

## figure 
out what ASVs are unambiguously a single species
```{r}
single_spp_seqs <- taxon_df %>% 
  group_by(qseqid) %>%
  add_tally(name = "n_taxa") %>%
  filter(n_taxa == 1)
```

## remove the single-species seqs from the dataframe and then rank the hits by % identity for the remaining seqs
```{r}
seq_id_diff <- taxon_df %>%
  anti_join(., single_spp_seqs) %>%  ## comment out this line if there are no single species ASVs
  select(-length) %>%
  group_by(qseqid, species, genus, family, order, class, phylum, kingdom) %>%
    mutate(seq_percID = max(perc_id)) %>%
    group_by(qseqid, species, genus, family, order, class, phylum, kingdom, seq_percID) %>%
  summarise(max(seq_percID)) %>% # take just the top hit for each taxon (for each sequence)
  select(-`max(seq_percID)`) %>%
  ungroup() %>%
  group_by(qseqid) %>%
      mutate(id_rank = rank(desc(seq_percID), ties.method = "min")) %>% # rank the taxonomic hits per sequence by % id
      mutate(top_perc = max(seq_percID)) %>% # designate the highest % id for the best taxonomic hit in each sequence (in some, but not all cases, this is 100%)   
      mutate(diff = top_perc - seq_percID) %>% # calculate the difference between the % identity of the top hit and each subsequent taxonomic hit
        arrange(diff)

seq_id_diff %>%
  filter(diff > 0)
```


## Figure out which differences are > 2% and eliminate those first
```{r}
# filter out any taxa that are >2% less matching identity than the top taxonomic hit for a given sequence
to_remove_low_perc_hits <- seq_id_diff %>%
  ungroup() %>%
  group_by(qseqid) %>%
  filter(diff > 1)

keepers <- seq_id_diff %>%
  anti_join(to_remove_low_perc_hits) #%>%
  #filter(seq_percID > 0.9799)
```

manually remove some odd entries from 'keepers'
```{r}
keepers_to_remove <- keepers %>%
  mutate(remove = ifelse(qseqid == "OTU_013" & species == "Lebbeus polaris", "yes", "no")) %>%
  mutate(remove = ifelse(qseqid == "OTU_043" & species == "Staphylococcus aureus", "yes", remove)) %>%
  mutate(remove = ifelse(qseqid == "OTU_048" & species == "eukaryote SCGC AAA071-J03", "yes", remove)) %>%
  mutate(remove = ifelse(qseqid == "OTU_048" & species == "Digitaria exilis", "yes", remove)) %>%
  mutate(remove = ifelse(qseqid == "OTU_050" & species == "Peridiniopsis cf. kevei diatom endosymbiont", "yes", remove)) %>%
  mutate(remove = ifelse(qseqid == "OTU_050" & species == "Peridiniopsis penardii diatom endosymbiont", "yes", remove)) %>%
  mutate(remove = ifelse(qseqid == "OTU_050" & species == "Unruhdinium penardii diatom endosymbiont", "yes", remove)) %>%
  mutate(remove = ifelse(qseqid == "OTU_061" & species == "Daptonema amphorum", "yes", remove)) %>%
  mutate(remove = ifelse(qseqid == "OTU_132" & species == "Selenicereus undatus", "yes", remove)) %>%
  mutate(remove = ifelse(qseqid == "OTU_007" & species == "Cyprinus carpio", "yes", remove)) %>%
  mutate(remove = ifelse(qseqid == "OTU_029" & species == "Phaeomyces dubiosus", "yes", remove)) %>%
  mutate(remove = ifelse(qseqid == "OTU_018" & species == "Phaeomyces dubiosus", "yes", remove)) %>%
  mutate(remove = ifelse(qseqid == "OTU_046" & species == "eukaryote SCGC AAA071-G16", "yes", remove)) %>%
  mutate(remove = ifelse(qseqid == "OTU_158" & species == "Halophytophthora vesicula", "yes", remove)) %>%
  mutate(remove = ifelse(qseqid == "OTU_011" & species == "Vigna unguiculata", "yes", remove)) %>%
  mutate(remove = ifelse(qseqid == "OTU_076" & species == "Chaenotheca xyloxena", "yes", remove)) %>%
  mutate(remove = ifelse(qseqid == "OTU_158" & species == "Sphaerephesia regularis", "yes", remove))

keepers <- keepers_to_remove %>%
  filter(remove == "no") %>%
  select(!remove)
```


## so now determine whether ASVs should be assigned to genus, family, order, etc. 
```{r}
singletons <- keepers %>%
  select(qseqid) %>%
  tally() %>%
  filter(n == 1)

# these are the seqs that now have only a single match
singleton_df <- singletons %>%
  left_join(keepers) %>%
  select(-n) %>%
  bind_rows(single_spp_seqs) %>% # combine the single spp data
  mutate(taxonomic_level = "species") %>%
  mutate(taxon = species)

## Genus-level matches
# remove the singletons from the bigger df 
single_genus <- keepers %>%
  anti_join(singleton_df)%>% # at best, these should be genus-level matches
  group_by(qseqid, genus) %>%
  tally() %>%
  ungroup() %>%
  group_by(qseqid) %>%
  tally() %>%
  filter(n == 1) %>% # seqs that match a single genus
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "genus") %>%
  mutate(taxon = genus)
  

## Family-level matches
single_family <- keepers %>%
  anti_join(singleton_df)%>%
  anti_join(single_genus) %>%
  group_by(qseqid, family) %>%
  tally() %>%
  ungroup() %>%
  group_by(qseqid) %>%
  tally() %>%
  filter(n == 1) %>% # seqs that match a single family
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "family") %>%
  mutate(taxon = family)


## Order-level matches
single_order <- keepers %>%
  anti_join(singleton_df)%>%
  anti_join(single_genus) %>%
  anti_join(single_family) %>%
  group_by(qseqid, order) %>%
  tally() %>%
  ungroup() %>%
  group_by(qseqid) %>%
  tally() %>%
  filter(n == 1) %>% # seqs that match a single order
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "order") %>%
  mutate(taxon = order)


## Class-level matches
single_class <- keepers %>%
  anti_join(singleton_df)%>%
  anti_join(single_genus) %>%
  anti_join(single_family) %>%
  anti_join(single_order) %>%
  group_by(qseqid, class) %>%
  tally() %>%
  ungroup() %>%
  group_by(qseqid) %>%
  tally() %>% 
  filter(n == 1) %>% # seqs that match a single class
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "class") %>%
  mutate(taxon = class)


## Phylum-level matches
single_phylum <- keepers %>%
  anti_join(singleton_df)%>%
  anti_join(single_genus) %>%
  anti_join(single_family) %>%
  anti_join(single_order) %>%
  anti_join(single_class) %>%
  group_by(qseqid, phylum) %>%
  tally() %>%
  ungroup() %>%
  group_by(qseqid) %>%
  tally() %>% 
  filter(n == 1) %>% # seqs that match a single phylum
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "phylum") %>%
  mutate(taxon = phylum)

# Kingdom level matches
single_kingdom <- keepers %>%
  anti_join(singleton_df)%>%
  anti_join(single_genus) %>%
  anti_join(single_family) %>%
  anti_join(single_order) %>%
  anti_join(single_class) %>%
  group_by(qseqid, kingdom) %>%
  tally() %>%
  ungroup() %>%
  group_by(qseqid) %>%
  tally() %>% 
  filter(n == 1) %>% # seqs that match a single kingdom
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "kingdom") %>%
  mutate(taxon = kingdom)
```

## modify the singleton_df to include the right variable headers
```{r}
single_spp <- singleton_df %>%
  select(-perc_id, -length, -sscinames, -n_taxa) %>%
  mutate(taxonomic_level = "species") %>%
  mutate(taxon = species)
```

## recombine the full data set now that the appropriate level of taxonomy has been determined

```{r}
sorted_tax_df <- bind_rows(single_kingdom, single_phylum, single_class, single_order, single_family, single_genus, single_spp)
```

## Create output taxonomy data frames
```{r}
uncollapsed_taxonomy <- sorted_tax_df %>%
  select(-top_perc, -id_rank) %>%
  unique() %>%
  arrange(qseqid)

# and then collapse that down to just a single taxon per ASV
collapsed_taxonomy <- uncollapsed_taxonomy %>%
  select(qseqid, taxon, taxonomic_level) %>%
  unique() %>%
  dplyr::rename(ASV = qseqid)

# make a few manual edits 
collapsed_taxonomy <- collapsed_taxonomy[-c(13,21,29,34,39),] %>%
  mutate(taxon = ifelse(ASV == "OTU_061", "Ascomycota/Basidiomycota", taxon)) %>%
    mutate(taxonomic_level = ifelse(ASV == "OTU_061", "phylum", taxonomic_level))
```


```{r}
write.csv(collapsed_taxonomy, "crab_18S_asv_taxonomy_blastn.csv")
```




