---
title: "determine taxonomic assignments for gadid ND1 asvs"
author: "Kimberly Ledger"
date: "2025-03-21"
output: github_document
---

- taxonomic assignment of gadid ND1 metabarcoding reads for Johanna' unknown cod 
- using dadasnake output from "20250311_gadid_id"
- using blastn output of ASVs (used ncbi nt db on sedna)  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-libraries}
library(tidyverse)
library(tidyr)
library(stringr)
rename <- dplyr::rename
```

## load output from taxonkit 
- this is the fasta seqs from the 20250130 run of dadasnake, and blastn using the ncbi nt database with 96% seq identity and 98% query coverage   

```{r load-output-from-taxonkit}
taxonomy <-read.delim("/home/kimberly.ledger/AFSC_eDNA/gadid_ID/data/blastn_taxlineage.txt", header = FALSE, na.strings=c(""," ","NA"))
head(taxonomy)
```


## clean up the header a bit
```{r}
# use the full taxonomy rather than the seq id to collapse identical entries
tax_df <- taxonomy %>%
  filter(V4 > 100) %>% # make sure all retained matches are >100 bp
  select(-V2, -V5, -V6, -V7, -V14, -V8, -V9, -V10, -V11, -V12, -V13, -V15, -V16) %>%  #remove unnecessary columns
  group_by(V1, V17) %>% # group by the sequence key and the full taxonomy to reduce duplicate entries
  unique() %>% # doing that reduced the number of entries
  rename(qseqid=V1, perc_id=V3, length=V4, taxonomy=V17) %>% #rename headers
  filter(perc_id > 98) %>% 
  filter(!is.na(taxonomy))
```

## calculate the number of ASV's with taxonomic assignment
```{r how-many-total-ASVs-w-taxonomy}
tax_df %>% 
  ungroup() %>%
  select(qseqid) %>%
  unique() %>%
  nrow
```

## reformat table 
```{r}
# formatting the taxonomy variables
taxon_df <- tax_df %>%
  filter(str_detect(taxonomy, ";")) %>%
  separate(taxonomy, into=c("kingdom", "phylum", "class", "order", "family", "genus", "species"), sep = ";") %>% 
  filter(!grepl("sp\\.", species, ignore.case = TRUE)) %>%
  filter(!grepl("cf\\.", species, ignore.case = TRUE)) %>%
  filter(!grepl("aff\\.", species, ignore.case = TRUE)) %>%
  filter(!grepl(" x ", species, ignore.case = TRUE)) %>%         ### remove hybrids and unknown species assignments
  filter(!grepl("synthetic", species, ignore.case = TRUE)) %>%
  filter(!grepl("uncultured", species, ignore.case = TRUE)) %>%
  filter(!grepl("\\d", species))
```

## what non-gadids are here? 
```{r}
not_cod <- taxon_df %>%
  filter(family != "Gadidae") 

unique(not_cod$species)
```
none. that's good. 

rename Gadus ogac to Gadus macro
```{r}
taxon_df <- taxon_df %>%
  mutate(species = ifelse(species == "Gadus ogac", "Gadus macrocephalus", species)) 
```

```{r}
tax_assignments <- taxon_df %>%
  select(qseqid, family, genus, species) %>%
  unique()
```

all asv's have unique species id. 

```{r}
#write.csv(tax_assignments, "/home/kimberly.ledger/AFSC_eDNA/gadid_ID/gadid_asv_taxonomic_assignments.csv")
```


now let's figure out which samples have which ASVs 

check sequence table outputs
```{r}
asv_table <- readRDS("/home/kimberly.ledger/AFSC_eDNA/gadid_ID/data/filtered.seqTab.RDS") %>%
  select(!Row.names)

#transpose 
asv_table <- data.frame(t(asv_table))

#set column names to be ASV# 
colnames(asv_table) <- asv_table["ASV",]

#remove row that has ASV#
asv_table <- asv_table[!rownames(asv_table) %in% c('ASV'), ]

#make sure reads are numbers
# Convert all character columns to numeric
for (col in names(asv_table)) {
  asv_table[[col]] <- as.numeric(asv_table[[col]])
}

asv_table$sample_ID <- rownames(asv_table)  #make make sample ID a column 
asv_table$sample_ID <- gsub("-", "_", asv_table$sample_ID) #illumina output changed "_" to "-"

#reorder columns and change NTC names to remove extra "_"
asv_table <- asv_table %>%
  select(sample_ID, everything())

asv_table[,-1] %>% summarise(across(everything(), sum))
```

```{r}
# make a variable for the first and last ASV column in the table
asv_first <- which(colnames(asv_table) == "ASV_26")  #this is the first asv column in this dataframe
asv_last <- ncol(asv_table)
```

pivot table longer 
```{r}
asv_table_long <- asv_table %>%
  pivot_longer(cols = c(asv_first:asv_last), names_to = "ASV", values_to = "reads") %>%
  mutate(reads = as.numeric(reads)) %>%
  mutate(reads = ifelse(is.na(reads), 0, reads)) %>%
  group_by(ASV) %>%
  mutate(reads_in_ASV = sum(reads))

length(unique(asv_table_long$ASV))
```

filter to remove ASVs with <100 reads across all samples 
```{r}
asv_table_filter1 <- asv_table_long %>%
  filter(reads_in_ASV > 99)
```

join to taxonomic assignments 
```{r}
taxon_table <- asv_table_filter1 %>%
  left_join(tax_assignments, by = c("ASV" = "qseqid")) %>%
  group_by(sample_ID, family, genus, species) %>%
  summarize(tot_reads = sum(reads))
```

calculate spp proportions 
```{r}
dominate_spp <- taxon_table %>%
  group_by(sample_ID) %>%
  mutate(reads_per_sample = sum(tot_reads),
         proportion = tot_reads/reads_per_sample) %>%
  filter(proportion > 0.9)
```

```{r}
write.csv(dominate_spp, "/home/kimberly.ledger/AFSC_eDNA/gadid_ID/sample_ID.csv")
```


